---
title: "Wasted Money: Toronto Police Budget and Crime Number"
subtitle: "Analyzing the changes in annual police budget and its effects
on the number of crimes within Toronto from 2020 to 2023"
author: 
  - Yisu Hou
thanks: "Code and data supporting this analysis are available at: 
https://github.com/YisuHou1/Toronto-Police-Budget-Statistics."
date: today
date-format: long
abstract: "An analysis of Toronto's annual police budget and crime counts
was conducted using R to determine the effectiveness of the budget in preventing
crime within the city from 2020 to 2023. It was found that the number of
reported crimes steadily increased while the rate of cleared crimes decreased
over time. Changes in the police budget, including raises from 2020 to 2022
and a reduction from 2022 to 2023, were uninfluential toward crime rates. The
finding that the police budget does not have positive or negative relationships
with crime insdicates that Toronto's current mode of crime prevention is
unproductive, and the police budget is potentially being wasted. It calls for
a need to reconsider how the safety of Toronto's citizens should be facilitated."
format: pdf
header-includes:
  - \usepackage{indentfirst}
  - \setlength{\parindent}{2em}
  - \setlength{\parskip}{0.5em}
  - \usepackage{float}
  - \usepackage{booktabs}
  - \usepackage{placeins}
number-sections: true
bibliography: references.bib
---


```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(knitr)
library(kableExtra)

data_bud <- read_csv(here::here("data/analysis_data/analysis_budget_data.csv"))
data_cri <- read_csv(here::here("data/analysis_data/analysis_crime_data.csv"))
summary_table_bud <- 
  read_csv(here::here("data/analysis_data/total_budget_data.csv"))
summary_table_cri <- 
  read_csv(here::here("data/analysis_data/total_crime_data.csv"))
```

# Introduction

In recently years, personal safety has developed into one
of the most prominent concerns among Toronto's citizens. @socialconnectedness
illustrated the recent rise in the number of violent crimes; the annual number
of shootings in 2020 was 462, almost double the average annual number from 2004
to 2015. The finding of @Putrik2019 constructs the connection between violence
and the lack of subjective safety, showing that people's perception of
higher rates of violent crimes, among other forms of crime, was the only factor
that had a positive correlation with feeling unsafe. The objective increase in
violent crimes and the subjective safety concerns compel individuals to question
whether the city's police department is fulfilling its purpose,
especially since the police consume a significant proportion of the city budget.

To date, no comprehensive study has addressed the relationship
between the police budget and the number of crimes in Toronto. However, such an
analysis is essential to assess the effectiveness of the police department in
preventing crime and protecting the safety of citizens. The study was conducted
with the purpose to fill this particular gap in the literature.

Relying on official data provided by Open Data Toronto @opendatatoronto and utilizing
R, a programming language created by @r, the annual police budget, the sum of
reported crimes, and the sum of cleared crimes were extracted for years 2020 to
2023. Using the data, it was identified that the number of crimes consistently
rose while the number of cleared crimes remained relatively stable, leading to
a reduction in the clearance ratio. The total annual budget of the police
department, calculated through subtracting revenues from the money spent, 
increased slightly from 2020 to 2022 but fell by over 40% in 2023. The drastic
change observed in the police budget data and the consistency of the crime number
data demonstrates Toronto police's lack of influence on crime. Thus, this paper
encourages a thorough analysis of why the investment toward Toronto's police
department was not translated into fewer crimes and increased safety.

The subsequent sections of this paper includes an introduction and justification
of the data source, an overview of the relevant variables, and a preliminary
analysis of the temporal trend between police budget and the number of crimes
in Toronto.


# Data {#sec-mydatasection}

Two separate datasets were extracted from Open Data Toronto @opendatatoronto, 
one recording all instances of expenditure and revenue since 2020, the other storing
the number of reported crimes in different categories and the corresponding
number of cleared crimes until the end of 2023. The summary statistics of the 
disaggregated data are located in the Appendix. They are not included in the body of
this paper because they are not used for analysis; the paper emphasizes aggregate
budgets and reported crime counts instead of individual paychecks and crime
categories. 

Open Data Toronto was selected to be the data source because it provides official
data from the City of Toronto, ensuring a decent level of credibility, variety,
and depth. The only alternative source of datasets on crime and police budget is 
the data portal of Toronto Police, and the datasets are identical to those in
Open Data Toronto. The measurements of the data, as provided by @opendatatoronto,
are not perfectly reliable. The police budget dataset is measured by addressing
approved budgets from Toronto Police Services Board and City Council and operation
expenses incurred. Expenses or revenues that were not officially reported were
excluded from the dataset. The dataset for crime reports and cleared crimes
relies on reports filed by officers, so undiscovered crimes are not
included, and unfounded reports are included. The biases in the measurements
increases the difficulty when interpreting trends. For example, if the number
of reported crimes increased over time, it might be due to an actual increase
in the number of crimes or an improvement in officers' ability to find crimes.

The cleaned, aggregated datasets are as follows:

```{r}
#| echo: false

# present the budget variables table
summary_table_bud %>%
  kable("latex", caption = "Total Police Budget Over Years {#tbl-bud}",
        col.names = c("Year", "Yearly Budget", 
                              "Change From Last", "Percent Change")) %>%
  kable_styling(latex_options = c("striped", "hold_position"))


```

\vspace{1em}

```{r}
#| echo: false

# present the crime variables table
summary_table_cri %>%
  kable("latex", caption = "Total Crime Reports Over Years {#tbl-cri}",
        col.names = c("Year", "Yearly Crimes", 
                              "Cleared Crimes", "Percent Cleared")) %>%
  kable_styling(latex_options = c("striped", "hold_position"))


```

Data cleaning and calculations were used to arrive at the variables in @tbl-bud
and @tbl-cri. "Year" was limited to 2020-2023 because the oldest data in the 
police budget dataset is 2020, and the most recent data in the crime report
dataset is 2023. "Yearly Budget" was calculated by summing all individual
expenses and revenues in the disaggregated dataset, which stores positive values for 
expenses and negative values for revenues. "Change From Last" was acquired by
subtracting the previous year's "Yearly Budget" from the current year, except
2020 because the previous year does not exist. "Percent Change" was calculated
by dividing the prevous year's "Yearly budget" by the current "Change From Last".
For @tbl-cri, "Yearly Crimes" and "Cleared Crimes" were calculated by summing
the reported cases and cleared cases of all crime categories in the disaggregated
dataset, respectively. The data cleaning proecss was aided by the R package
tidyverse @tidyverse.


```{r}
#| label: fig-bud
#| fig-cap: "Total Police Budget Over Time"
#| echo: false

# graphs to show change in police budget and crime rates over time
ggplot(data = summary_table_bud, aes(x = Year, y = Yearly_Budget)) + 
  geom_bar(stat = "identity", fill = "skyblue") + 
  labs(
    x = "Year",
    y = "Yearly Budget ($)"
  )

```

\vspace{1em}

```{r}
#| label: fig-cri
#| fig-cap: "Total Reported Crimes and Cleared Crimes Over Time"
#| echo: false

# plot for crime and cleared crime count over time
ggplot(summary_table_cri, aes(x = factor(Year))) + 
  # Plot total crime count as a full bar
  geom_bar(aes(y = Yearly_Crimes, fill = "Total Crimes"), stat = "identity",
           position = "stack") +
  
  # Overlay cleared crimes as a portion of the total crimes
  geom_bar(aes(y = Cleared_Crimes, fill = "Cleared Crimes"), 
           stat = "identity", position = "stack") +
  
  # Custom colors
  scale_fill_manual(values = c("Total Crimes" = "tomato", 
                               "Cleared Crimes" = "skyblue")) +
  
  # Labels and theme
  labs(
    x = "Year",
    y = "Number of Crimes",
    fill = "Crime Type"
  ) +
  theme_minimal()

```


\vspace{1em}

\FloatBarrier

@tbl-bud and @fig-bud shows minor increases in the police budget from 2020 to 2022, 
and a near-half reduction in 2023. @fig-cri illustrates consistent increases in
total crimes from 2021 to 2023, while the cleared cases remained relatively
static throughout four years. Thus, it is unsurprising to observe that in
@tbl-cri, the rate of cleared crimes decreased steadily, demonstrating the
persistent deterioration of the police's ability to prevent crime, regardless of
the budget. It can be argued that the amount of resources possessed by Toronto
Police does not impact its productivity, and it is necessary to reconsider how
the city budget should be allocated. Taking this another step further, the fact
that altering the police budget does not impact the number of crimes indicates
that our policing is fundamentally ineffective, or that the laws, even enforced
through policing, does not prevent crime. Therefore, it is imperative to assess 
whether the current modes of policing and lawmaking have real impacts on
criminals' decision-making.



\newpage

\appendix

# Appendix {.unnumbered}

The appendix is left blank for now. I need to add raw tables here and update
tests. The plots are created with @knitr and @kextra

\newpage

# References {.unnumbered}


